---
title: "Getting which Weather Event the most effects in population health and economy from NOAA Data"
output:
  pdf_document: default
  html_document: default
---

# Getting which Weather Event the most effects in population health and economy

This analysis focuses in discovering which kind of weather event the most effects in population health and economy in the USA between 1950 and 2011. The data is  recollected from the U.S. National Oceanic and Atmospheric Administration’s [NOAA](https://www.noaa.gov/) service. 

## Data Processing

Obviously, the first step is preparing the environment. It means downloading the data and loading needed packages.  
```{r, message= FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)

urlData <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
pathData <- "data.csv.bz2"
if (!file.exists(pathData)) {
  download.file(urlData, destfile = "data.csv.bz2")
}
```

Later, it is only considered 14 variables based on the relation with the questions for the project and the available data in the data set (there exists a lot of NA values in those variables).  
```{r}
noaa <- read.csv(pathData, na.strings = "")
names_c <- c("BGN_DATE", "COUNTYNAME", "STATE", "EVTYPE", "LENGTH", "WIDTH", "F", "MAG", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
noaa <- noaa[, names_c]

```

```{r}
naPROP <- sum(is.na(noaa$PROPDMGEXP))
noaa$PROPDMGEXP <- ifelse(is.na(noaa$PROPDMGEXP), '0',noaa$PROPDMGEXP)
naCROP <- sum(is.na(noaa$CROPDMGEXP))
noaa$CROPDMGEXP <- ifelse(is.na(noaa$CROPDMGEXP), '0',noaa$CROPDMGEXP)
```
There are `r naPROP` NA values in the variable property damage exponation.  
There are `r naCROP` NA values in the variable crop damage exponation.   

So, it is neccesary fixing with 0 values since it will not generated any problems later.  

## Results

#### Question 1: Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?  
For this question, The dplyer package is used to grouping based on type of event and summarise the total sum of fatalities and injuries. Additionally, observations with zero values in both variables (fatalities and injuries) are removed.  
```{r}
noaa_by_evtype <- noaa %>% group_by(EVTYPE) %>% summarise(tot_fatalities = sum(FATALITIES), tot_injuries = sum(INJURIES)) %>% filter(!(tot_fatalities==0 & tot_injuries == 0))
```

Over this new data set we order based on total injuries to see which event type causes the most damage in population health.  
For injuries in population:  
```{r}
index_max_injuries <- which.max(noaa_by_evtype$tot_injuries)
event_injury <- noaa_by_evtype[index_max_injuries,1]
noaa_by_evtype[order(noaa_by_evtype$tot_injuries, decreasing = TRUE), c(1, 3)][1:10,]
```
It shows clearly that `r event_injury` is the weather event which generates the most damage causing injuries.  
Similarly, for fatalities in population:  
```{r}
index_max_fatalities <- which.max(noaa_by_evtype$tot_fatalities)
event_fatality <- noaa_by_evtype[index_max_fatalities,]
noaa_by_evtype[order(noaa_by_evtype$tot_fatalities, decreasing = TRUE), c(1, 2)][1:10,]
```
It shows clearly that `r event_fatality` is the weather event which generates the most damage causing fatalities.    

**So based on evidences `r event_fatality` generates the most health damage between wheather events.**


#### Question 2: Across the United States, which types of events have the greatest economic consequences?

For the second question we need to clean a little more the data. Basically, we have to convert all the economic information to the same unit. For that reason, *convertWIthBase* function helps to convert all units to the same format.
```{r}

convertWithBase <- function(v) {
  
  val <- as.numeric(v[1])
  b <- v[2]
  c <- substr(b, 1, 1)
  numb <- c("0","1","2","3","4","5","6","7","8","9")
  if (c == '-' || c == '?' || c == '+') {
    val
  } else if (c %in% numb) {
    val * 10 ** as.numeric(b)    
  } else if (c == 'h' || c == 'H') {
    val * 100
  } else if (c == 'k' || c == 'K') {
    val * 1000
  } else if (c == 'm' || c == 'M') {
    val * 1000000
  } else if (c == 'b' || c == 'B') {
    val * 1000000000
  } else {
    val
  }
}

noaa$CROPDMGTOT <- apply(data.frame(crop=noaa$CROPDMG, exp=noaa$CROPDMGEXP), MARGIN = 1, FUN = convertWithBase)

noaa$PROPDMGTOT <- apply(data.frame(crop=noaa$PROPDMG, exp=noaa$PROPDMGEXP), MARGIN = 1, FUN = convertWithBase)


```

Later, with all data with the same economic unit, it is necessary summarize by damage in properties and crops. With the intention to get more visibility I applied a filter to get the events with the most economic effect.
```{r}
noaa_by_evtype_econ <- noaa %>% group_by(EVTYPE) %>% summarise(tot_prop = sum(PROPDMGTOT), tot_crop = sum(CROPDMGTOT)) %>% filter(!(tot_prop==0 & tot_crop == 0))
```

One first table show the most 5 weather events which causing the most damage in properties. 
```{r}
event_prop <- noaa_by_evtype_econ[which.max(noaa_by_evtype_econ$tot_prop),1]
noaa_prop <- noaa_by_evtype_econ[order(noaa_by_evtype_econ$tot_prop, decreasing = T),c(1, 2)][1:5,]
noaa_prop
```
In similar way, the table show the most 5 weather events which causing the most damage in crops. 
```{r}

event_crop <-noaa_by_evtype_econ[which.max(noaa_by_evtype_econ$tot_crop), 1]
noaa_crop <- noaa_by_evtype_econ[order(noaa_by_evtype_econ$tot_crop, decreasing = T), c(1, 3)][1:7,]
noaa_crop
```

Finally, the table show the most 5 weather events which causing the most damage in crops and properties in total. 
```{r}
noaa_by_evtype_econ$tot_dam <- noaa_by_evtype_econ$tot_prop + noaa_by_evtype_econ$tot_crop
noaa_dam <- noaa_by_evtype_econ[order(noaa_by_evtype_econ$tot_dam, decreasing = T), c(1, 4)][1:6,]
noaa_dam
```

For a better visualization, one plot with the 3 criterios is showed.

```{r}

plot_prop <- ggplot(noaa_prop, aes(x=EVTYPE, y=tot_prop,fill=EVTYPE, label = round(tot_prop,2))) +
    geom_bar(stat="identity") +
    xlab("Event Type") +
    ylab("Total damage in properties") +
    ggtitle("Damage in properties based on event type")+
    geom_label(aes(fill = EVTYPE),colour = "white", fontface = "bold") +
    theme(legend.position="none")

plot_crop <- ggplot(noaa_crop, aes(x=EVTYPE, y=tot_crop,fill=EVTYPE, label = round(tot_crop,2))) +
    geom_bar(stat="identity") +
    xlab("Event Type") +
    ylab("Total damage in crops") +
    ggtitle("Damage in crops based on event type")+
    geom_label(aes(fill = EVTYPE),colour = "white", fontface = "bold") +
    theme(legend.position="none")

plot_dam <- ggplot(noaa_dam, aes(x=EVTYPE, y=tot_dam,fill=EVTYPE, label = round(tot_dam,2))) +
    geom_bar(stat="identity") +
    xlab("Event Type") +
    ylab("Total damage") +
    ggtitle("Damage based on event type")+
    geom_label(aes(fill = EVTYPE),colour = "white", fontface = "bold") +
    theme(legend.position="none")

fig <- function(width, heigth){
     options(repr.plot.width = width, repr.plot.height = heigth)
}
fig(1000, 4000)
options(repr.plot.width = 14, repr.plot.height = 8)
grid.arrange(plot_prop, plot_crop, plot_dam, nrow=3)
```


For this case, it shows that *TORNADO* is the weather event which generates the most damage in economic.


